---
output:
  BiocStyle::html_document
---

<!---
Because we split the analysis pipeline in different independent files,
to speed up processing it, here in the setup block we load libraries and
objects that were loaded or produced in the previously processed file,
and which are necessary in this file.
--->

```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(SummarizedExperiment)
library(edgeR)
library(geneplotter)

opts_chunk$set(cache=TRUE,
               cache.path="cache/file2",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")

se.unpaired <- readRDS(file.path("rawCounts", "seLUSC.rds"))
dg.unpaired <- readRDS(file.path("results", "dge.rds"))

se <- readRDS(file.path("results", "se.paired.rds"))
dge <- readRDS(file.path("results", "dge.paired.rds"))

se.filt.unnorm <- readRDS(file.path("results", "se.paired.filt.unnorm.rds"))
dge.filt.unnorm <- readRDS(file.path("results", "dge.paired.filt.unnorm.rds"))
se.filt <- readRDS(file.path("results", "se.paired.filt.rds"))
dge.filt <- readRDS(file.path("results", "dge.paired.filt.rds"))
```

# Differential expression

We perform a simple examination of expression changes and their associated p-values
using the R/Bioconductor package [sva](http://bioconductor.org/packages/sva).

In order to use the method, a matrix of the model, including adjustment variables, and a null model which only includes the adjustment variables have to be created. In our case there are no factors to adjust, so we give ~1 as the intercept term in the model.

```{r, message=FALSE}
library(sva)
mod <- model.matrix(~type + 1, data = colData(se.filt))
mod0 <- model.matrix(~1, data = colData(se.filt))
sv <- sva(assays(se.filt)$logCPM, mod, mod0)
pv <- f.pvalue(assays(se.filt)$logCPM, mod, mod0)
sum(p.adjust(pv, method="fdr") < 0.01)
```

There are `r sum(p.adjust(pv, method="fdr") < 0.01)` genes changing significantly their expression at FDR < 1%. In Figure \@ref(fig:pdist) below we show the distribution of the resulting p-values. This distribution should be uniform with a peak at low p-values for the truly DE genes, assuming that most genes are not differentially expressed.

```{r pdist, echo=FALSE, out.width="600px", fig.cap="Distribution of raw p-values for an F-test on every gene between tumor and normal samples."}
hist(pv, main="", las=1)
```

The `sva()` function is able to estimate surrogate variables, which have continuous values that can be used to correct and adjust unmeasured and unwanted effects.

```{r}
sv <- sva(assays(se.filt)$logCPM, mod, mod0)
sv$n
```

The SVA algorithm has found `r sv$n` surrogate variables. Let's use them to assess againt the extent of differential expression this time adjusting for these surrogate variables.

```{r}
modsv <- cbind(mod, sv$sv)
mod0sv <- cbind(mod0, sv$sv)
pvsv <- f.pvalue(assays(se.filt)$logCPM, modsv, mod0sv)
sum(p.adjust(pvsv, method="fdr") < 0.01)
```

We have increased the number of changing genes to `r sum(p.adjust(pvsv, method="fdr") < 0.01)`. Figure \@ref(fig:psvdist) shows the resulting distribution of p-values.

```{r psvdist, echo=FALSE, out.width="600px", fig.cap="Distribution of raw p-values for an F-test on every gene between tumor and normal samples, adjusting for surrogate variables estimated with SVA."}
hist(pvsv, main="", las=1)
```

## Session information

```{r, message=FALSE}
sessionInfo()
```
