---
title: "Analysis of a TCGA RNA-seq data set on lung squamous cell carcinoma"
author:
- name: Carbonell Gamón, Gabriel
  affiliation:
  - &id Master Programme on Bioinformatics for Health Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: gabriel.carbonell01@estudiant.upf.edu
- name: Matabacas Enebral, Aleix
  affiliation: *id
  email: aleix.matabacas01@estudiant.upf.edu
- name: McKitrick, Austin
  affiliation: *id
  email: austin.mckitrick01@estudiant.upf.edu
date: "`r format(Sys.time(), '%B %e, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: bibliography.bib
---

# Introduction

Among all cancer types, lung cancer is the most lethal among both males and females. It is responsible for 28% of cancer-related mortality. The life expectation for those who have lung cancer are very poor. It is usually diagnosed in an advanced phase, and the patients only have a 16% survival rate after 5 years of their diagnosis (@nature11404).

The Cancer Genome Atlas (TCGA) has comprehensively profiled this type of cancer in a patient cohort (@rahman2015alternative). Here, we analyze the expression profiles of those patients, using a pipeline based on the R/Bioconductor software package `r Biocpkg("Rsubread")`.

This document is written in [R markdown](https://rmarkdown.rstudio.com) and
should be processed using R and you need to install the packages
`r CRANpkg("knitr")` and `r CRANpkg("markdown")`. Moreover, it using the official style for Bioconductor vignettes facilitated by the Bioconductor package `r Biocpkg("BiocStyle")`. It is compiled from different files following the instructions included in the `makefile`.


```{r echo=FALSE, results='asis'}
cat(readLines("QAanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("DEanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("FEanalysis.knit.md"), sep="\n")
```

# References
