---
title: "Analysis of a TCGA RNA-seq data set on lung squamous cell carcinoma"
author:
- name: Gabriel Carbonell
  affiliation:
  - &id Master Programme on Bioinformatics for Health Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: gabriel.carbonell01@estudiant.upf.edu
- name: Aleix Matabacas
  affiliation: *id
  email: aleix.matabacas01@estudiant.upf.edu
- name: Austin McKitric
  affiliation: *id
  email: austin.mckitric01@estudiant.upf.edu
date: "`r format(Sys.time(), '%B %e, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: bibliography.bib
---

# Introduction

Lung cancer is the most lethal one in comparison with other cancer types, in both genders. Is the responsibility of the 28\% in the all cancer mortality. The expects for the ones who have this kind of cancer diagnosed are very poor, Usually is diagnosed in an advanced phase and the patients only have a 16\% of survival rate after 5 years of their diagnosis.

The Cancer Genome Atlas (TCGA) has comprehensively profiled this type of cancer in a patient cohort. Here we analyze the expression profiles of those
patients, using a pipeline based on the R/Bioconductor software package `r Biocpkg("Rsubread")`.

This document is written in [R markdown](https://rmarkdown.rstudio.com) and
should be processed using R and you need to install the packages
`r CRANpkg("knitr")` and `r CRANpkg("markdown")`. Moreover, it using the official style for Bioconductor vignettes facilitated by the Bioconductor package `r Biocpkg("BiocStyle")`. 

   ```
   $ make
   ```
```{r echo=FALSE, results='asis'}
cat(readLines("QAanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("DEanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("FEanalysis.knit.md"), sep="\n")
```
